## Quickstart

Before getting into the details, we'll first go over a quick example to provide a feel for what can be done with Trelliscope.  This example is adapted from the quick start example in the [datadr](http://tessera.io/docs-datadr) documentation, but with a specific focus on Trelliscope.

### Package installation

First, we need to install the necessary components, `datadr` and `trelliscope`.  These are R packages that we install from github.

```{r quickstart_install, eval=FALSE, echo=TRUE}
# install packages (one time only)
install.packages("devtools") # if not installed
devtools::install_github("tesseradata/datadr")
devtools::install_github("tesseradata/trelliscope")
```

Our example is based on a small dataset that we can handle in a local R session, and therefore we only need to have these two packages installed.  For support of more scalable back ends like Hadoop when dealing with larger data sets, see the [quickstart section](http://tessera.io/#quickstart) on the Tessera website.

We will use as an example a data set consisting of the median list and sold price of homes in the United States, aggregated by county and month from 2008 to early 2014, reported from [Zillow](http://www.zillow.com) and obtained from [quandl](https://www.quandl.com).  A pre-processed version of this data is available in a package called `housingData`, which we will use.  To install this package:

```{r quickstart_install2, eval=FALSE, echo=TRUE}
devtools::install_github("hafen/housingData")
```

### Environment setup

Now we load the packages and look at the housing data:

```{r quickstart_loadpackages, eval=TRUE, echo=TRUE, message=FALSE}
# load packages
library(housingData)
library(datadr)
library(trelliscope)

# look at housing data
str(housing)
```

We see that we have a data frame with the information we discussed, in addition to the number of units sold.

### Setting up a visualization database

We create many plots throughout the course of analysis, and with Trelliscope, we can store these in a "visualization database" (VDB), which is a directory on our computer where all of the information about our display artifacts is stored.  Typically we will set up a single VDB for each project we are working on.  To initialize and connect to a VDB, we call the `vdbConn()` function with the path where our VDB is located (or where we would like it to be located), and optionally give it a name.

```{r quickstart_vdb, eval=TRUE, echo=TRUE, results="hide", message=FALSE}
# connect to a "visualization database"
conn <- vdbConn("vdb", name = "tesseraTutorial")
```

This connects to a directory called `"vdb"` relative to our current working directory.  The first time you do this it will ask to make sure you want to create the directory.  R holds this connection in its global options so that subsequent calls will know where to put things without explicitly specifying the connection each time.

### Visualization by county and state

Trelliscope allows us to visualize large data sets in detail.  We do this by splitting the data into meaningful subsets and applying a visualization to each subset, and then interactively viewing the panels of the display.

An interesting thing to look at with the housing data is the median list and sold price over time by county and state.  To split the data in this way, we use the `divide()` function from the `datadr` package.  It is recommended to have some familiarity with the [datadr](http://tessera.io/docs-datadr) package.

```{r quickstart_loaddata, eval=TRUE, echo=FALSE, results="hide", message=FALSE, purl=FALSE}
# byCounty <- divide(housing,
#    by = c("county", "state"), update = TRUE)
# save(byCounty, file = "housing/byCounty.Rdata")
load("_ignore/housing/byCounty.Rdata")
```


```{r quickstart_divide, eval=FALSE, echo=TRUE}
# divide housing data by county and state
byCounty <- divide(housing,
   by = c("county", "state"))
```

Our `byCounty` object is now a distributed data frame (ddf), which is simply a data frame split into chunks of key-value pairs.  The key defines the split, and the value is the data frame for that split.  We can see some of its attributes by printing the object:

```{r quickstart_divide_print, eval=TRUE, echo=TRUE}
# look at byCounty object
byCounty
```

And we can look at one of the subsets:

```{r quickstart_divide_print2, eval=TRUE, echo=TRUE}
# look at a subset of byCounty
byCounty[[1]]
```

The key tells us that this is Abbeville county in South Carolina, and the value is the price data for this county.

### Creating a panel function

To create a Trelliscope display, we need to first provide a *panel* function, which specifies what to plot for each subset.  It takes as input either a key-value pair or just a value, depending on whether the function has two arguments or one.

For example, here is a panel function that takes a value and creates a lattice `xyplot` of list and sold price over time:

```{r quickstart_panel, eval=TRUE, echo=TRUE}
# create a panel function of list and sold price vs. time
timePanel <- function(x)
   xyplot(medListPriceSqft + medSoldPriceSqft ~ time,
      data = x, auto.key = TRUE, ylab = "Price / Sq. Ft.")
```

Note that you can use most any R plot command here (base R plots, lattice, ggplot, rCharts, ggvis).

 test it on a subset:
```{r quickstart_panel_test, eval=TRUE, echo=TRUE, fig.height=4}
# test function on a subset
timePanel(byCounty[[20]]$value)
```

Great!

### Creating a cognostics function

Another thing we can do is specify a *cognostics* function for each subset.  A cognostic is a metric that tells us an interesting attribute about a subset of data, and we can use cognostics to have more worthwhile interactions with all of the panels in the display.  A cognostic function needs to return a list of metrics:

```{r quickstart_cog, eval=TRUE, echo=TRUE}
# create a cognostics function of metrics of interest
priceCog <- function(x) {
   zillowString <- gsub(" ", "-", do.call(paste, getSplitVars(x)))
   list(
      slope = cog(coef(lm(medListPriceSqft ~ time, data = x))[2],
         desc = "list price slope"),
      meanList = cogMean(x$medListPriceSqft),
      meanSold = cogMean(x$medSoldPriceSqft),
      nObs = cog(length(which(!is.na(x$medListPriceSqft))),
         desc = "number of non-NA list prices"),
      zillowHref = cogHref(
         sprintf("http://www.zillow.com/homes/%s_rb/", zillowString),
         desc = "zillow link")
   )
}
```

We use the `cog()` function to wrap our metrics so that we can provide a description for the cognostic, and we also employ special cognostics functions `cogMean()` and `cogRange()` to compute mean and range with a default description.

We should test the cognostics function on a subset:

```{r quickstart_cog_test, eval=TRUE, echo=TRUE}
# test cognostics function on a subset
priceCog(byCounty[[1]]$value)
```

### Making the display

Now we can create a Trelliscope display by sending our data, our panel function, and our cognostics function to `makeDisplay()`:

```{r quickstart_makedisplay, eval=FALSE, echo=TRUE}
# create the display and add to vdb
makeDisplay(byCounty,
   name = "list_sold_vs_time_quickstart",
   desc = "List and sold price over time",
   panelFn = timePanel,
   cogFn = priceCog,
   width = 400, height = 400,
   lims = list(x = "same"))
```

This creates a new entry in our visualization database and stores all of the appropriate information for the Trelliscope viewer to know how to construct the panels.

If you have been dutifully following along with this example in your own R console, you can now view the display with the following:

```{r name, eval=FALSE, echo=TRUE}
view()
```

And select the display with the name "list_sold_vs_time_quickstart".

If you have not been following along but are wondering what that `view()` command did, you can visit <a href="http://hafen.shinyapps.io/trelliscopeTutorial/" target="_blank">here</a> for an online version.  You will find a list of displays to choose from, of which the one with the name `list_sold_vs_time_quickstart` is the one we just created.  This brings up the point that you can share your Trelliscope displays online -- more about that as well as how to use the viewer will be covered in the Trelliscope tutorial -- but feel free to play around with the viewer and see what you can discover.

This covers the basics of `trelliscope`.  Hopefully you now feel comfortable enough to dive into the rest of the tutorial.
